---
title: "Standard Report"
author: "Paul Hewson"
date: "2025-05-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
 
Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
You need a working pandoc installation on your computer (this should be taken care of if you have an RStudio installation)


```{r load_data_and_packages, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(targets)
tar_load(summary_data)
library(dplyr)
library(ggplot2)
```

## Road casualty data

We can present a visual display of the number of road casualties as follows:

```{r pressure, echo=FALSE}
summary_data %>%
  ggplot(aes(x = casualty_severity, y = Casualties, 
             fill = casualty_severity)) +
  geom_bar(stat = "identity") + 
  theme_bw() + 
  xlab("Severity of Casualty") + 
  ylab("Number of reported and sub-sampled casualties") + 
  labs(title = "Number of reported pedestrian road casualties in 2023",
       subtitle = "All fatalities, 10% of serious and 1% of slight casualties",
       caption = "Source: STATs19, data.gov.uk")
```

Do note that there has been a lot of sub-sampling. All fatally injured road pedestrian casualties are included in the data, but 10% of serious and 1% of slight casualties were selected at random.


```{r eval=FALSE, message=FALSE}
library(targets)
# Read back the two evaluation targets:
rf <- tar_read(rf_eval)
svm <- tar_read(svm_eval)

# Print AUCs
cat("**Random Forest AUC:**", round(rf$auc, 3), "\n\n")
cat("**SVM (Radial) AUC:**", round(svm$auc, 3), "\n\n")

# Print confusion matrices
knitr::kable(rf$confusion$table, 
             caption = "Random Forest Confusion Matrix", 
             format = "markdown")
knitr::kable(svm$confusion$table,
             caption = "SVM (Radial) Confusion Matrix",
             format = "markdown")
```


That will render nicely-formatted tables of your predicted vs. true class counts and report the AUCs inline.

---

## 2. Plot feature‐importance for the Random Forest

Caret’s `varImp()` will extract importances from your `ranger` model:

```markdown
```{r eval=FALSE, fig.height=6, fig.width=8}
library(caret)
# Get variable importance
vi <- varImp(tar_read(rf_fit))$importance
vi$variable <- rownames(vi)

# Plot the top 10
library(ggplot2)
ggplot(vi %>% arrange(desc(Overall)) %>% slice(1:10),
       aes(x = reorder(variable, Overall), y = Overall)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Predictors by Variable Importance (Random Forest)",
    x = "", y = "Importance"
  ) +
  theme_minimal()
```

---

## 3. Discuss why RF > SVM here

Below your tables & plot, add a prose section:

> **Why does Random Forest outperform the SVM?**  
>
> 1. **Non-linear interactions & high-dimensional splits**  
>    RF can capture complex interactions among features (e.g. time × lighting) via recursive binary splits.  A single radial SVM, by contrast, may struggle to find a kernel hyperplane that cleanly separates three classes in this mixed numeric/categorical space.  
>
> 2. **Ensemble stability vs. SVM sensitivity**  
>    The bagged‐trees ensemble in RF reduces variance and is robust to outliers and noisy predictors.  SVM tuning (cost/gamma) can be brittle with overlapping classes—hence its much lower recall on “Serious” cases (≈47% vs. ≈87% in RF).  
>
> 3. **Native multiclass handling**  
>    Ranger handles multiclass natively; the SVM via `svmRadial` in Caret uses one-vs-one behind the scenes, which may introduce inconsistencies when classes overlap.  
>
> Taken together, these strengths let RF achieve ~0.98 multiclass AUC and ~90% balanced accuracy, compared to only ~0.79 AUC and ~60% overall accuracy for the SVM.

---

Putting it all together in your `Report.Rmd` will give an analytical narrative:  
- **“Here are the confusion tables & AUCs…”**  
- **“Here’s which predictors matter most…”**  
- **“Here’s why RF is the clear winner on these data.”**

That completes Task 1’s final “communicate & interpret” requirement.
